Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-10 numerical-amr.tex
Minor edits to overview, AMR and Zeus sections.

Unlike moving mesh methods \citep{1995ApJS..100..269P,1995ApJS...97..231G} or  methods that subdivide individual cells \citep{Adjerid}, Berger \& Collela's AMR (also referred to as \emph{structured} AMR) utilizes an adaptive hierarchy of grid patches at varying levels of resolution.  The SAMR method described here is similar to \citet{Berger89}.  A more complete discussion of the history and development of this type of algorithm is available in \citet{Neeman96}.

@@

The basic control algorithm looks very much like any used in a grid-based code of a single resolution, and can be written schematically as:
\begin{verbatim}
InitializeHierarchy
While (Time < StopTime)
begin
  dt = ComputeTimeStep(0)
  EvolveLevel(0, dt)
  Time = Time + dt
  CheckForOutput(Time)
end
\end{verbatim}
The AMR control algorithm is contained within the recursive
EvolveLevel algorithm, which takes the level on which to operate as an
argument.  This looks something like the following pseudo-code:
\begin{verbatim}
EvolveLevel(level)
begin
  SetBoundaryValues
  while (Time < ParentTime)
  begin
    dt = ComputeTimeStep(level)
    SolveHydroEquations(dt)
    SolveOtherEquations(dt)
    Time = Time + dt
    SetBoundaryValues
    EvolveLevel(level+1, dt)
    FluxCorrection
    Projection
    RebuildHierarchy(level+1)
  end
end
\end{verbatim}

@@

Given the grid and its boundary conditions, the process of solving
equations~(\ref{eq:mass})--(\ref{eq:total_energy}) can be detached
from the rest of the SAMR technique, so that it is possible to
experiment with a number of different methods.  The restrictions are
that the method should be flux-conservative, or at least produce an
accurate estimate of the flux of conserved quantities at cell
interfaces.  This is required in the flux-correction step (see
\ref{sec:flux_correction}).  There is also the issue of centering,
which refers to the position of physical quantities within a cell (the
primary types are cell-centered, face-centered, and node-centered). In this description we 
will employ the direct Eulerian version of the piecewise
parabolic method \citep{1984JCoPh..54..174C,1995CoPhC..89..149B}, which
assumes cell-centering for all quantities (actually they are
cell-averages).  However, it is also possible to use other schemes,
which employ different centering for different variables
\citep[e.g.,][]{Stone92a}.  The interpolation step, described in
Section~\ref{sec:interpolation}, must correctly interpolate with the
selected centering.

@@

We first find the smallest region (or ``proto-subgrid") that covers all of the refined cells on a grid.  We compute an efficiency $\epsilon$ for this grid, defined as the ratio of flagged cells to total cells.  If this efficiency is sufficiently large (typically over 30\%), or if the grid is small enough (typically less than $4^d$ coarse cells) then we adopt the region as a new grid and move on to the next region (if any).  If the region is not acceptable, then we split it into two using the following algorithm.  First, we determine the longest dimension and compute the signature along that dimension.  The signature is found by summing the number of refined cells with the same index along the longest dimension.  Edges are typically indicated by inflection points in this signature, so we calculate the second derivative using a simple three-point finite difference estimator and look for the maximum in the second derivative, with ties going to the index closest to the center.  If no inflection points are found along the longest axis, the next-longest axis is searched.  If no inflection points are found, the longest axis is cut in half.  The original region is then split into two new regions, or proto-subgrids.  This process is then re-applied to each of the new regions.

>>>>

%Unlike moving mesh methods \citep{1995ApJS..100..269P,1995ApJS...97..231G} or  methods that subdivide individual cells \citep{Adjerid}, Berger \& Collela's AMR (also referred to as \emph{structured} AMR) utilizes an adaptive hierarchy of grid patches at varying levels of resolution.  The SAMR method described here is similar to \citet{Berger89}.  A more complete discussion of the history and development of this type of algorithm is available in \citet{Neeman96}.

@@

The basic control algorithm looks very much like any used in a grid-based code of a single resolution, and can be written schematically as:
\begin{verbatim}
InitializeHierarchy
While (Time < StopTime)
   dt = ComputeTimeStep(0)
   EvolveLevel(0, dt)
   Time = Time + dt
  CheckForOutput(Time)
\end{verbatim}
The AMR control algorithm is contained within the recursive
EvolveLevel algorithm, which takes the level on which to operate as an
argument.  This looks something like the following pseudo-code:
\begin{verbatim}
EvolveLevel(level)
   SetBoundaryValues
   while (Time < ParentTime)
      dt = ComputeTimeStep(level)
      SolveHydroEquations(dt)
      SolveOtherEquations(dt)
      Time = Time + dt
      SetBoundaryValues
      EvolveLevel(level+1, dt)
      FluxCorrection
      Projection
      RebuildHierarchy(level+1)
\end{verbatim}

@@

Given the grid and its boundary conditions, the process of solving
equations~(\ref{eq:mass})--(\ref{eq:total_energy}) can be detached
from the rest of the SAMR technique, so that it is possible to
experiment with a number of different methods.  The restrictions are
that the method should be flux-conservative, or at least produce an
accurate estimate of the flux of conserved quantities at cell
interfaces.  This is required in the flux-correction step (see
\ref{sec:flux_correction}).  There is also the issue of centering,
which refers to the position of physical quantities within a cell (the
primary types are cell-centered, face-centered, and node-centered). 
%In this description we will employ the direct Eulerian version of the piecewise
%parabolic method \citep{1984JCoPh..54..174C,1995CoPhC..89..149B}, which
%assumes cell-centering for all quantities (actually they are
%cell-averages).  
%However, it is also possible to use other schemes,
%which employ different centering for different variables
%\citep[e.g.,][]{Stone92a}.  
The interpolation step, described in
Section~\ref{sec:interpolation}, must correctly interpolate with the
selected centering.

@@

We first find the smallest region (or ``proto-subgrid") that covers all of the refined cells on a grid.  We compute an efficiency $\epsilon$ for this grid, defined as the ratio of flagged cells to total cells.  If this efficiency is sufficiently large (typically over 30\%), or if the grid is small enough, then we adopt the region as a new grid and move on to the next region (if any).  If the region is not acceptable, then we split it into two using the following algorithm.  First, we determine the longest dimension and compute the signature along that dimension.  The signature is found by summing the number of refined cells with the same index along the longest dimension.  Edges are typically indicated by inflection points in this signature, so we calculate the second derivative using a simple three-point finite difference estimator and look for the maximum in the second derivative, with ties going to the index closest to the center.  If no inflection points are found along the longest axis, the next-longest axis is searched.  If no inflection points are found, the longest axis is cut in half.  The original region is then split into two new regions, or proto-subgrids.  This process is then re-applied to each of the new regions.

### 2013-06-10 numerical-zeus.tex
Minor edits to overview, AMR and Zeus sections.

A limitation of a technique that uses an artificial viscosity is that, while the correct Rankine-Hugoniot jump conditions are achieved, shocks are broadened over 3-4 mesh cells and are thus not treated as true discontinuities. This may cause unphysical pre-heating of gas upstream of the shock wave, as discussed in \citet{1994ApJ...429..434A}.  On the other hand, it is much more robust than PPM and is easy to add additional physics.  We also note that this method solves only the internal energy equation rather than total energy, so the ``dual energy formulation'' discussed in Section~\ref{sec.hydro.ppm} is unnecessary.

>>>>

A limitation of a technique that uses an artificial viscosity is that, while the correct Rankine-Hugoniot jump conditions are achieved, shocks are broadened over 3-4 mesh cells. This may cause unphysical pre-heating of gas upstream of the shock wave, as discussed in \citet{1994ApJ...429..434A}.  On the other hand, it is much more robust than PPM and is easy to add additional physics.  We also note that this method solves only the internal energy equation rather than total energy, so the dual energy formulation discussed in Section~\ref{sec.hydro.ppm} is unnecessary.

### 2013-06-10 overview.tex
Minor edits to overview, AMR and Zeus sections.

The code can include either a homogeneous radiation background (with
possible local self-shielding corrections), or evolve an inhomogeneous
radiation field either by directly solving the radiative transfer
equation along rays or by solving a set of moment equations derived
from the radiative transfer equation. In comoving coordinates
\citep[e.g.,][]{Gnedin97} the radiative transfer equation reads
%
\begin{equation}
  \label{eq:rteqn}
  \frac{1}{c} \; \frac{\partial I_\nu}{\partial t} + 
  \frac{a_{\rm em}}{a} \, \hat{n} \cdot \grad I_\nu -
  \frac{H}{c} \; \left( \nu \frac{\partial I_\nu}{\partial \nu} -
  3 I_\nu \right) = -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
Here $I_\nu \equiv I(\nu, \mathbf{x}, \Omega, t)$ is specific
intensity of the radiation, with dimensions of energy per time $t$ per
solid angle $\Omega$ per unit area per frequency $\nu$.  The second
term represents the propagation of radiation, where the factor $a_{\rm
  em}/a$ accounts for cosmic expansion since the time of emission.
The third term describes both the cosmological redshift and dilution
of radiation.  On the right hand side, the first term captures
absorption ($\kappa_\nu \equiv \kappa_\nu(\mathbf{x},\nu,t)$), and the
second term emission ($j_\nu \equiv j_\nu(\mathbf{x},\nu,t)$ from any
sources of radiation (whether point or diffuse).

@@

We begin with the direct-Eulerian PPM implementation.  This is an
explicit, higher-order-accurate version of Godunov's method for ideal
gas dynamics with third order accurate piecewise parabolic monotonic
interpolation and a nonlinear Riemann solver for shock capturing.  It
advances the hydrodynamic equations in the following steps:
%\begin{enumerate}
% \item 
(i) Construct monotonic parabolic interpolation of cell average data,
for each fluid quantity;
% \item 
(ii) Compute interface states by averaging the parabola over the
domain of dependence for each interface;
% \item 
(iii) Use interface data to solve the Riemann problem;
 %\item 
(iv) Difference the interface fluxes to update the cell average
quantities.
%\end{enumerate}

>>>>

The code can include either a homogeneous radiation background, 
or evolve an inhomogeneous
radiation field either by directly solving the radiative transfer
equation along rays or by solving a set of moment equations derived
from the radiative transfer equation. In comoving coordinates
\citep[e.g.,][]{Gnedin97} the radiative transfer equation reads
%
\begin{equation}
  \label{eq:rteqn}
  \frac{1}{c} \; \frac{\partial I_\nu}{\partial t} + 
  \frac{a_{\rm em}}{a} \, \hat{n} \cdot \grad I_\nu -
  \frac{H}{c} \; \left( \nu \frac{\partial I_\nu}{\partial \nu} -
  3 I_\nu \right) = -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
Here $I_\nu \equiv I(\nu, \mathbf{x}, \Omega, t)$ is specific
intensity of the radiation, with dimensions of energy per time $t$ per
solid angle $\Omega$ per unit area per frequency $\nu$.  The second
term represents the propagation of radiation, where the factor $a_{\rm
  em}/a$ accounts for cosmic expansion since the time of emission.
The third term describes both the cosmological redshift and dilution
of radiation.  On the right hand side, the first term captures
absorption ($\kappa_\nu \equiv \kappa_\nu(\mathbf{x},\nu,t)$), and the
second term emission ($j_\nu \equiv j_\nu(\mathbf{x},\nu,t)$ from any
sources of radiation (whether point or diffuse).

@@

We begin with the direct-Eulerian PPM implementation.  This is an
explicit, higher-order-accurate version of Godunov's method for ideal
gas dynamics with the spatially third order accurate piecewise parabolic monotonic
interpolation and a nonlinear Riemann solver for shock capturing.  It
advances the hydrodynamic equations in the following steps:
%\begin{enumerate}
% \item 
(i) Construct monotonic parabolic interpolation of cell average data,
for each fluid quantity;
% \item 
(ii) Compute interface states by averaging the parabola over the
domain of dependence for each interface;
% \item 
(iii) Use interface data to solve the Riemann problem;
 %\item 
(iv) Difference the interface fluxes to update the cell average
quantities.
%\end{enumerate}
